Polynomial phase estimation by least squares phase unwrapping

### 2013-07-06 paper.tex
minor

The paper is organised in the following way. Section~\ref{sec:lattice-theory} describes some required concepts from lattice theory.  In Section~\ref{sec:ident_aliasing} we describe the identifiable region that was derived in~\cite{McKilliam2009IndentifiabliltyAliasingPolyphase}.  These identifiability results are required in order to understand the statistical properties of the LSU estimator.  Section~\ref{sec:least-squar-unwr} describes the LSU estimator and states a theorem asserting the estimator to be strongly consistent and asymptotically normally distributed under some mild assumptions on the noise process $\{X_n\}$. The proof of strong consistency is given in the appendix.  Space restrictions make it impossible to give a proof of asymptotic normality in this paper.  A complete proof of this is available in~\cite{McKilliam_arxiv_pps_unwrapping_2012}.  Section~\ref{sec:simulations} describes the results of Monte Carlo simulations that compare the LSU estimator with existing estimators including the HAF, PHAF and CPF.  These simulations agree with the asymptotic properties.

@@

A \term{lattice},  $\Lambda$, is a discrete subset of points in $\reals^n$ for which
\[
   \Lambda = \{\xbf = \Bbf\ubf \mid \ubf \in \ints^d \}
\]
where $\Bbf \in \reals^{n \times d}$ has rank $d$ and is called the generator matrix.  If $n = d$ the lattice is said to be of full rank.  Lattices are discrete Abelian groups under vector addition.  They are subgroups of the Euclidean group $\reals^n$.  Lattices naturally give rise to tessellations of $\reals^n$ by the specification of a set of coset representatives for the quotient $\reals^n / \Lambda$.  One choice for a set of coset representatives is a fundamental parallelepiped, the parallelepiped generated by the columns of a generator matrix.  Another choice is based on the Voronoi cell, those points from $\reals^n$ nearest (with respect to the Euclidean norm in this paper) to the lattice point at the origin.  It is always possible to construct a rectangular set of representatives, as the next proposition will show.  We will use these rectangular regions to describe the aliasing properties of polynomial phase signals in Section~\ref{sec:ident_aliasing}.  These rectangular regions will be important for the derivation of the asymptotic properties of the LSU estimator.
% in Section~\ref{sec:least-squar-unwr}.

@@

\begin{theorem} \label{thm:asymp_proof} 
Let $\widehat{\mubf}$ be defined by~\eqref{eq:hatmubfLSUdefn} and put $\widehat{\lambdabf}_N = \dealias(\tilde{\mubf} - \widehat{\mubf})$.  Denote the elements of $\widehat{\lambdabf}_N$ by $\widehat{\lambda}_{0,N}, \dots, \widehat{\lambda}_{m,N}$.  Suppose $\Phi_1, \dots, \Phi_N$ are independent and identically distributed with zero intrinsic mean, intrinsic variance $\sigma^2$, and pdf $f$, then: 
\begin{enumerate}
\item (Strong consistency) $N^k \widehat{\lambda}_{k,N}$ converges almost surely to $0$ as $N\rightarrow\infty$ for all $k = 0, 1, \dots, m$.
\item (Asymptotic normality) If $f(\fracpart{x})$ is continuous at $x = -\nicefrac{1}{2}$ and $f(-\nicefrac{1}{2}) < 1$ then the distribution of
\[
\left[
\begin{array}
[c]{cccc}%
\sqrt{N} \widehat{\lambda}_{0,N} & N \sqrt{N}\widehat{\lambda}_{1,N}  & \dots & N^m\sqrt{N} \widehat{\lambda}_{m,N}
\end{array}
\right]^\prime
\]
converges to the normal with zero mean and covariance matrix
\[
\frac{\sigma^2}{\left(1-f( -\nicefrac{1}{2}) \right)^{2}} \Cbf^{-1},
\]
where $\Cbf$ is the $(m+1)\times (m+1)$ Hilbert matrix with elements $C_{ik} = 1/(i + k + 1)$ for $i,k \in \{0, 1, \dots, m\}$.
\end{enumerate}
\end{theorem}

A proof of the first part of this theorem (strong consistency) is given in Appendix~\ref{sec:strongconstproof}.  Space restrictions make it impossible prove asymptotic normality in this paper.  A complete proof of normality is available in~\cite{McKilliam_arxiv_pps_unwrapping_2012}.  Proofs for the case of $m=0$ were given in~\cite{McKilliam_mean_dir_est_sq_arc_length2010} and $m=1$ in~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009}.  The proof here takes a similar approach, but requires new techniques.  %Before giving the proof of strong consistency we make the following remarks  
A statement similar to Theorem~\ref{thm:asymp_proof} without a complete proof was given in~\cite{McKilliam_polyphase_est_icassp_2011}.  Here, we give a proof.  The results here are also more general than in~\cite{McKilliam_polyphase_est_icassp_2011}, allowing for a wider class of noise distributions.

@@

This section describes the results of Monte-Carlo simulations with the least squares unwrapping (LSU) estimator, the high order ambiguity function (HAF) estimator~\cite{Peleg_DPT_1995}, the product high order ambiguity function (PHAF) estimator~\cite{Barbarossa_PHAF_1998}, the cubic phase function (CPF) estimator~\cite{Oshea_cpf_2004} and the estimator that results from combining the HAF and CPF~\cite{Djurovic_haf_cpf_2012}.  The CPF estimator only applies to polynomial phase signals of order $m=3$ and when the number of observations $N$ is odd.  In all simulations the unknown amplitude is $\rho = 1$ and $N = 199$.  The $X_1, \dots, X_N$ are pseudorandomly generated independent and identically distributed circularly symmetric complex Gaussian with variance $\expect\abs{X_1}^2 = \sigma_c^2$.  The corresponding signal to noise ratio (SNR) is $\rho^2 \sigma_c^{-2} = \sigma_c^{-2}$.  The number of replications of each Monte-Carlo experiment is $T = 2000$ to obtain estimates $\widehat{\mubf}_1, \dots, \widehat{\mubf}_T$ and the corresponding dealiased errors $\widehat{\lambdabf}_t = \dealias(\tilde{\mubf} - \widehat{\mubf}_t)$ are computed.  The sample mean square error (MSE) of the $k$th coefficient is computed according to $\tfrac{1}{T}\sum_{t=1}^T \widehat{\lambda}_{k,t}^2$ where $\widehat{\lambda}_{k,t}$ is the $k$th element of $\widehat{\lambdabf}_t$.  In all simulations four \emph{lag sets} are used for the PHAF estimator~\cite{Barbarossa_PHAF_1998} and the LSU estimator is approximated using the $K$-best method described in Section~\ref{sec:lattice-theory}.  The maximum number of branches allowed is $K=20N$.

@@

The comparatively small set of coefficients suitable for the HAF, PHAF and CPF estimators has been studied previously~\cite{Angeby_PPS_aliasing_2000,Djurovic_aliasing_2012,McKilliam2009IndentifiabliltyAliasingPolyphase,Zhou_lag_diverity_PPT_1997,Xia_dynamic_range_pps_2001,Morelande_bayes_unwrapping_2009_tsp} and can be a critical limitation for some applications, for example, interference suppression in the presence of a jammer~\cite{Djukanovic:2011:PMN:1950994.1951252,Djurovic_aliasing_2012}.  Zhou and Wang~\cite{Zhou_lag_diverity_PPT_1997} describe a method based on the extended Euclidean algorithm that enables the HAF estimator to operate on the entire identifiable region $B$.  Unfortunately this comes at the price of statistical accuracy.  Figure~\ref{plot:m3bigzw} shows the performance of the Zhou and Wang estimator when the coefficients are $\tilde{\mubf} = \left[ \tfrac{1}{4}, \tfrac{1}{4}, \tfrac{1}{8}, \tfrac{1}{24}  \right]$, the same as in Figure~\ref{plot:m3big}.  The threshold for the estimator is approximately~\unit[29]{dB}.  There does not appear to exist (at least in the literature) a similar method for increasing the range of coefficients for the PHAF and CPF estimators.

@@

Tables~\ref{tablm1},~\ref{tablm3} and~\ref{tablm5} show the values of SNR at which the estimators reach a threshold for polynomial phase signals of order $m=1,3$ and $5$.  The values in the tables are approximated using Monte-Carlo simulation with $X_1, \dots, X_N$ pseudorandomly generated independent and identically distributed circularly symmetric complex Gaussian.  When $m=3$ the true coefficients are as in Figure~\ref{plot:m3small} and when $m=5$ the true coefficients are as in Figure~\ref{plot:m5} so that the coefficients are inside the range acceptable for the HAF, CPF and CPF-HAF estimators.  When $m=1$ (single frequency estimation) the LSU estimator can be computed exactly using the algorithm from~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009}.  In this case the value at which the LSU estimator reaches a threshold is similar to the maximum likelihood periodogram estimator~\cite{Quinn2001,Hannan1973,Rife1974}.  This property was observed in~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009}.  Both the LSU and periodogram estimators are strongly consistent, and so, the threshold values can be made arbitrarily small by choosing $N$ large.  This is indicated in the bottom row of Table~\ref{tablm1} by the $-\infty$ symbols.  When $m>1$ the LSU estimator can be exactly computed using the sphere decoder~\cite{Agrell2002,Viterbo_sphere_decoder_1999} in a reasonable amount of time when $N < 60$.  When $N > 60$ we use the approximate $K$-best algorithm.  We observe that the $K$-best algorithm produces estimates with the same threshold as the exactly computed LSU estimator when $N = 19$ and $49$.  As $N\rightarrow\infty$ the asymptotic threshold of the HAF estimator has been studied in~\cite{Porat_asympt_HAF_DPT_1996} and these values are indicated in the bottom rows of tables~\ref{tablm3} and~\ref{tablm5}.  By Theorem~\ref{thm:asymp_proof} the LSU estimator is strongly consistent and the threshold can be made as small as desired by choosing $N$ large.  Whether the LSU estimator computed approximately using the $K$-best algorithm is strongly consistent is an open theoretical question.  That the $K$-best algorithm provides a close approximation of the exact LSU estimator for small $N$ is an encouraging observation.

@@

\begin{table}[t]
\centering
\begin{tabular}{lrrrrr}
 & LSU & LSU ($K$-best) & CPF & HAF & ZW \\ \toprule
$N=19$ & 10 & 10 & 8 & 11 & 31 \\ 
$N=49$ & 7 & 7 & 5 & 9 & 29 \\ 
$N=99$ & & 6 & 3 & 5 & 27 \\ 
$N=199$ & & 3 & 1 & 3 & 23 \\ 
$N=299$ & & 2 & -1 &  & \\ 
$N\to\infty$ & $-\infty$ & & & 3.77 & \\ \bottomrule
\end{tabular}
\caption{Estimated SNR threshold in decibels with $m=3$}
\label{tablm3}
\end{table}

\begin{table}[t]
\centering
\begin{tabular}{lrrrrr}
 & LSU & LSU ($K$-best) & CPF-HAF & HAF & ZW \\ \toprule
$N=19$ & 22  & 22 & 17 & 23 & $>100$ \\ 
$N=49$ & 17 & 17 & 14 & 21 & $>100$  \\ 
$N=99$ &  & 15 &  12 &  19 & $>100$ \\ 
$N=199$ & & 12 & 10 & 18 & 103 \\ 
$N=299$ &  &  &  & &  \\ 
$N\to\infty$ & $-\infty$ &  &  & 16.6 &  \\ \bottomrule
\end{tabular}
\caption{Estimated SNR threshold in decibels with $m=5$}
\label{tablm5}
\end{table}

@@

This paper has considered the estimation of the coefficients of a noisy polynomial phase signal by least squares phase unwrapping (LSU). The LSU estimator was shown to be strongly consistent and asymptotically normally distributed under mild conditions on the distribution of the noise.  The estimator can be computed by finding a nearest lattice point in a lattice. Polynomial time nearest point algorithms for these lattices exist~\cite[Sec 4.3]{McKilliam2010thesis}, but these algorithms are not fast in practice.  We instead make use of the approximate $K$-best algorithm~\cite{Baro_list_sphere_dec_2003,Zhan2006_K_best_sphere_decoder,Mondal_Kbest_again_2009}.  This algorithm results in an accurate estimator and can be computed in a reasonable amount of time for moderately large numbers of observations $N$.

>>>>

The paper is organised in the following way. Section~\ref{sec:lattice-theory} describes some required concepts from lattice theory and Section~\ref{sec:ident_aliasing} describes the identifiable region that was derived in~\cite{McKilliam2009IndentifiabliltyAliasingPolyphase}.  These identifiability results are required in order to understand the statistical properties of the LSU estimator.  Section~\ref{sec:least-squar-unwr} describes the LSU estimator and states a theorem asserting the estimator to be strongly consistent and asymptotically normally distributed under some mild assumptions on the noise process $\{X_n\}$. The proof of strong consistency is given in the appendix.  Space restrictions make it impossible to give a proof of asymptotic normality in this paper.  A complete proof of this is available in~\cite{McKilliam_arxiv_pps_unwrapping_2012}.  Section~\ref{sec:simulations} describes the results of Monte Carlo simulations that compare the LSU estimator with existing estimators including the HAF, PHAF and CPF.  These simulations agree with the asymptotic properties.

@@

A \term{lattice},  $\Lambda$, is a discrete subset of points in $\reals^n$ for which
\[
   \Lambda = \{\xbf = \Bbf\ubf \mid \ubf \in \ints^d \}
\]
where $\Bbf \in \reals^{n \times d}$ has rank $d$ and is called the generator matrix.  If $n = d$ the lattice is said to be of full rank.  Lattices are discrete Abelian groups under vector addition.  They are subgroups of the Euclidean group $\reals^n$.  Lattices naturally give rise to tessellations of $\reals^n$ by the specification of a set of coset representatives for the quotient $\reals^n / \Lambda$.  One choice for a set of coset representatives is a fundamental parallelepiped, the parallelepiped generated by the columns of a generator matrix.  Another choice is based on the Voronoi cell, those points from $\reals^n$ nearest (with respect to the Euclidean norm in this paper) to the lattice point at the origin.  It is always possible to construct a rectangular set of representatives, as the next proposition will show.  These rectangular regions are used to describe the aliasing properties of polynomial phase signals in Section~\ref{sec:ident_aliasing}.  These rectangular regions will be important for the derivation of the asymptotic properties of the LSU estimator.
% in Section~\ref{sec:least-squar-unwr}.

@@

\begin{theorem} \label{thm:asymp_proof} 
Let $\widehat{\mubf}$ be defined by~\eqref{eq:hatmubfLSUdefn} and put $\widehat{\lambdabf}_N = \dealias(\tilde{\mubf} - \widehat{\mubf})$.  Denote the elements of $\widehat{\lambdabf}_N$ by $\widehat{\lambda}_{0,N}, \dots, \widehat{\lambda}_{m,N}$.  Suppose $\Phi_1, \dots, \Phi_N$ are independent and identically distributed with zero intrinsic mean, intrinsic variance $\sigma^2$, and pdf $f$, then: 
\begin{enumerate}
\item (Strong consistency) $N^k \widehat{\lambda}_{k,N}$ converges almost surely to $0$ as $N\rightarrow\infty$ for all $k = 0, 1, \dots, m$.
\item (Asymptotic normality) If $f(\fracpart{x})$ is continuous at $x = -\nicefrac{1}{2}$ and $f(-\nicefrac{1}{2}) < 1$ then the distribution of
\[
\left[
\begin{array}
[c]{cccc}%
\sqrt{N} \widehat{\lambda}_{0,N} & N \sqrt{N}\widehat{\lambda}_{1,N}  & \dots & N^m\sqrt{N} \widehat{\lambda}_{m,N}
\end{array}
\right]^\prime
\]
converges to the normal with zero mean and covariance matrix
\[
\frac{\sigma^2}{\left(1-f( -\nicefrac{1}{2}) \right)^{2}} \Cbf^{-1},
\]
where $\Cbf$ is the $m+1$ by $m+1$ Hilbert matrix with elements $C_{ik} = 1/(i + k + 1)$ for $i,k \in \{0, 1, \dots, m\}$.
\end{enumerate}
\end{theorem}

A proof of the first part of this theorem (strong consistency) is given in Appendix~\ref{sec:strongconstproof}.  Space restrictions make it impossible to prove asymptotic normality in this paper.  A complete proof of normality is available in~\cite{McKilliam_arxiv_pps_unwrapping_2012}.  Proofs for the case of $m=0$ were given in~\cite{McKilliam_mean_dir_est_sq_arc_length2010} and $m=1$ in~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009}.  The proof here takes a similar approach, but requires new techniques.  %Before giving the proof of strong consistency we make the following remarks  
A statement similar to Theorem~\ref{thm:asymp_proof} without a complete proof was given in~\cite{McKilliam_polyphase_est_icassp_2011}.  Here, we give a proof.  The results here are also more general than in~\cite{McKilliam_polyphase_est_icassp_2011}, allowing for a wider class of noise distributions.

@@

This section describes the results of Monte-Carlo simulations with the least squares unwrapping (LSU) estimator, the high order ambiguity function (HAF) estimator~\cite{Peleg_DPT_1995}, the product high order ambiguity function (PHAF) estimator~\cite{Barbarossa_PHAF_1998}, the cubic phase function (CPF) estimator~\cite{Oshea_cpf_2004} and the estimator that results from combining the HAF and CPF~\cite{Djurovic_haf_cpf_2012}.  The CPF estimator only applies to polynomial phase signals of order $m=3$ and when the number of observations $N$ is odd.  In all simulations the unknown amplitude is $\rho = 1$ and $N = 199$.  The $X_1, \dots, X_N$ are pseudorandomly generated independent and identically distributed circularly symmetric complex Gaussian with variance $\expect\abs{X_1}^2 = \sigma_c^2$.  The corresponding signal to noise ratio (SNR) is $\rho^2 \sigma_c^{-2} = \sigma_c^{-2}$.  The number of replications of each Monte-Carlo experiment is $T = 2000$ to obtain estimates $\widehat{\mubf}_1, \dots, \widehat{\mubf}_T$ and the corresponding dealiased errors $\widehat{\lambdabf}_t = \dealias(\tilde{\mubf} - \widehat{\mubf}_t)$ are computed.  The sample mean square error (MSE) of the $k$th coefficient is computed according to $\tfrac{1}{T}\sum_{t=1}^T \widehat{\lambda}_{k,t}^2$ where $\widehat{\lambda}_{k,t}$ is the $k$th element of $\widehat{\lambdabf}_t$.  In all simulations four \emph{lag sets} are used for the PHAF estimator~\cite{Barbarossa_PHAF_1998} and the LSU estimator is approximated using the $K$-best method described in Section~\ref{sec:lattice-theory}.  The maximum number of branches allowed is $K=20N$.  The arithmetic complexity of the LSU estimator is $O(N^3\log N)$ in this case.

@@

The comparatively small set of coefficients suitable for the HAF, PHAF and CPF estimators has been studied previously~\cite{Angeby_PPS_aliasing_2000,Djurovic_aliasing_2012,McKilliam2009IndentifiabliltyAliasingPolyphase,Zhou_lag_diverity_PPT_1997,Xia_dynamic_range_pps_2001,Morelande_bayes_unwrapping_2009_tsp} and can be a critical limitation for some applications, for example, interference suppression in the presence of a jammer~\cite{Djukanovic:2011:PMN:1950994.1951252,Djurovic_aliasing_2012}.  Zhou and Wang~\cite{Zhou_lag_diverity_PPT_1997} describe a method based on the extended Euclidean algorithm that enables the HAF estimator to operate on the entire identifiable region $B$.  Unfortunately this comes at the price of statistical accuracy.  Figure~\ref{plot:m3bigzw} shows the performance of the Zhou and Wang estimator when the coefficients are $\tilde{\mubf} = \left[ \tfrac{1}{4}, \tfrac{1}{4}, \tfrac{1}{8}, \tfrac{1}{24}  \right]$, the same as in Figure~\ref{plot:m3big}.  The threshold for the estimator is approximately~\unit[28]{dB}.  There does not appear to exist (at least in the literature) a similar method for increasing the range of coefficients for the PHAF and CPF estimators.

@@

Tables~\ref{tablm1},~\ref{tablm3} and~\ref{tablm5} show the values of SNR at which the estimators reach a threshold for polynomial phase signals of order $m=1,3$ and $5$.  The values in the tables are approximated using Monte-Carlo simulation with $X_1, \dots, X_N$ pseudorandomly generated independent and identically distributed circularly symmetric complex Gaussian.  When $m=3$ the true coefficients are as in Figure~\ref{plot:m3small} and when $m=5$ the true coefficients are as in Figure~\ref{plot:m5} so that the coefficients are inside the range acceptable for the HAF, CPF and CPF-HAF estimators.  When $m=1$ (single frequency estimation) the LSU estimator can be computed exactly using the algorithm from~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009}.  In this case the value at which the LSU estimator reaches a threshold is similar to the maximum likelihood periodogram estimator~\cite{Quinn2001,Hannan1973,Rife1974}.  This property was observed in~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009}.  Both the LSU and periodogram estimators are strongly consistent, and so, the threshold values can be made arbitrarily small by choosing $N$ large.  This is indicated in the bottom row of Table~\ref{tablm1} by the $-\infty$ symbols.  When $m>1$ the LSU estimator can be exactly computed using the sphere decoder~\cite{Agrell2002,Viterbo_sphere_decoder_1999} in a reasonable amount of time when $N < 60$.  When $N > 60$ we use the approximate $K$-best algorithm.  We observe that the $K$-best algorithm produces estimates with the same threshold as the exactly computed LSU estimator when $N = 19$ and $49$.  As $N\rightarrow\infty$ the asymptotic threshold of the HAF estimator has been studied in~\cite{Porat_asympt_HAF_DPT_1996} and these values are indicated in the bottom rows of Tables~\ref{tablm3} and~\ref{tablm5}.  By Theorem~\ref{thm:asymp_proof} the LSU estimator is strongly consistent and the threshold can be made as small as desired by choosing $N$ large.  Whether the LSU estimator computed approximately using the $K$-best algorithm is strongly consistent is an open theoretical question.  That the $K$-best algorithm provides a close approximation of the exact LSU estimator for small $N$ is an encouraging observation.

@@

\begin{table}[t]
\centering
\begin{tabular}{lrrrrr}
 & LSU & LSU ($K$-best) & CPF & HAF & ZW \\ \toprule
$N=19$ & 10 & 10 & 8 & 11 & 31 \\ 
$N=49$ & 7 & 7 & 5 & 9 & 29 \\ 
$N=99$ & & 6 & 3 & 7 & 29 \\ 
$N=199$ & & 3 & 1 & 5 & 28 \\ 
$N=299$ & & 1 & -1 & 5 & 27 \\ 
$N\to\infty$ & $-\infty$ & & & 3.77 & \\ \bottomrule
\end{tabular}
\caption{Estimated SNR threshold in decibels with $m=3$}
\label{tablm3}
\end{table}

\begin{table}[t]
\centering
\begin{tabular}{lrrrrr}
 & LSU & LSU ($K$-best) & CPF-HAF & HAF & ZW \\ \toprule
$N=19$ & 22  & 22 & 17 & 23 & $>100$ \\ 
$N=49$ & 17 & 17 & 14 & 21 & $>100$  \\ 
$N=99$ &  & 15 &  12 &  19 & $>100$ \\ 
$N=199$ & & 12 & 10 & 18 & $>100$ \\ 
$N=299$ &  & 11 & 9 & 17 & $>100$ \\ 
$N\to\infty$ & $-\infty$ &  &  & 16.6 &  \\ \bottomrule
\end{tabular}
\caption{Estimated SNR threshold in decibels with $m=5$}
\label{tablm5}
\end{table}

@@

This paper has considered the estimation of the coefficients of a noisy polynomial phase signal by least squares phase unwrapping (LSU). The LSU estimator was shown to be strongly consistent and asymptotically normally distributed under mild conditions on the distribution of the noise.  The estimator can be computed by finding a nearest point in a lattice. Polynomial time nearest point algorithms for these lattices exist~\cite[Sec 4.3]{McKilliam2010thesis}, but these algorithms are not fast in practice.  We instead make use of the approximate $K$-best algorithm~\cite{Baro_list_sphere_dec_2003,Zhan2006_K_best_sphere_decoder,Mondal_Kbest_again_2009}.  This algorithm results in an accurate estimator and can be computed in a reasonable amount of time for moderately large numbers of observations $N$.
